Spring haul-out behavior of seals in the Bering and Chukchi seas

### 2024-02-17 analysis/paper/London_HauloutBeahvior.Rmd
ver hoef and crawford comments addressed

  breeding, molting, pupping, and resting. In the Arctic, many of these activities
  occur in spring (April -- June) as sea ice begins to melt and retreat northward.
  Rapid acceleration of climate change in Arctic ecosystems is therefore of
  concern as the quantity and quality of suitable habitat is forecast to decrease.
  Improved estimates of seal population abundance are needed to properly monitor
  the impacts of these changes over time. In this paper, we use hourly
  percent-dry data from satellite-linked bio-loggers deployed
  between 2005 and 2021 to quantify the proportion of seals hauled out on ice.
  This information is needed to accurately estimate abundance from aerial survey counts of
  ice-associated seals (i.e., to correct for the proportion of animals that
  are in the water while surveys are conducted). In addition to providing
  essential data for survey 'availability' calculations, our analysis also provides 
  insights into the seasonal timing and environmental

@@

  snow lairs, they were not included in this analysis. Using generalized linear mixed
  pseudo-models to properly account for temporal autocorrelation, we fit models
  with covariates of interest (e.g., day-of-year, solar hour, age-sex class, wind
  speed, barometric pressure, temperature, precipitation) to examine their ability
  to explain variation in haul-out probability. We found evidence for
  strong diel and within-season patterns in haul-out behavior, as well as
  strong weather effects (particularly wind and temperature). In general, seals
  were more likely to haul out on ice in the middle of the day and when wind speed
  was low and temperatures were higher. Haul-out probability increased through
  March and April, peaking in May and early June before declining again. The
  timing and frequency of haul-out events also varied based on species and age-sex
  class. For ribbon and spotted seals, models with year effects were highly
  supported, indicating that the timing and magnitude of haul-out behavior varied
  among years. However, we did not find broad evidence that haul-out timing was linked
  to annual sea ice extent. Our analysis emphasizes the importance of accounting
  for seasonal and temporal variation in haul-out behavior, as well as associated
  environmental covariates, when interpreting the number of seals counted in
  aerial surveys.

@@

Statutory requirements (e.g., United States Marine Mammal Protection Act (MMPA),
United States Endangered Species Act (ESA)) for timely estimates of population
abundance and trend of these species mean improved aerial survey effort is
needed -- and, must be paired with improved knowledge of haul-out behavior to
ensure appropriate survey design, robust methods, and accurate estimates.
Several studies have contributed estimates of the distribution and abundance of
ice-associated seal species in the Arctic using aerial surveys (e.g.,
@bengtson2005a, @conn2014a, and @verhoef2014a) and more recent efforts
have significantly expanded on previous survey effort. Such abundance studies
are conducted over very large areas and estimation of absolute abundance
requires making inference about numerous issues affecting the observation of
seals on ice. These include availability (only seals on ice are available to
be counted), detection probability (observers or automated detection systems may
miss some seals on ice), species misclassification, and possible
disturbance of seals by aircraft [@conn2014a; @verhoef2014a]. Refining these
inferences will improve the accuracy of abundance estimates and, hopefully,
allow credible predictions about the effects of climate disruptions on the
abundance and distribution of Arctic seal populations.

How ice-associated seals use sea ice as a haul-out platform varies between species.
Ribbon seals (_Histriophoca fasciata_) haul out of the water almost exclusively
on sea ice and are mostly pelagic outside the spring pupping, breeding, and
molting season [@boveng2018a]. Although spotted (_Phoca largha_) and bearded
(_Erignathus barbatus_) seals rest on coastal features, they strongly prefer sea
ice as a haul-out platform during the spring and early summer [@frost2018].
Ringed seals (_Phoca hispida_) --- not included in this study --- haul out on
sea-ice but also within snow lairs during winter and spring.

@@

In this study, we used 16 years of bio-logging data to investigate the haul-out
behavior of bearded, ribbon, and spotted seals in the Bering and Chukchi seas.
Our goals were threefold. First, we wished to establish baseline estimates for
the chronology of haul-out behavior in the critical spring season for each
species across different age and sex classes. Second, we sought to refine
estimates of haul-out availability corrections for aerial surveys in order to
improve estimates of seal abundance. Previously estimated availability
correction factors (e.g., @bengtson2005a, @conn2014a, and @verhoef2014a)
accounted for variables such as the time-of-day and day-of-year, but did not
investigate the impact of meteorological variables. Such variables have been
shown to influence walrus haul-out behavior [@udevitz2009] and we expect weather
conditions to also influence seal haul-out behavior and including them within
the model framework will benefit our estimates of seal availability during
aerial surveys. Third, we aimed to assess the annual variability in haul-out
timing and possible linkage to changes in the extent of seasonal sea ice between
2005 and 2020. Our work extends the scope of previous haul-out analyses,
includes the influence of meteorological variability, and investigates the
potential impact of changing sea-ice extent on the behavior of these species.

@@

For this study we used haul-out behavior data and location estimates from
bio-loggers deployed on bearded, ribbon, and spotted seals in the Bering,
Chukchi, and western Beaufort seas by multiple organizations as part of
collaborative investigations from 2005 through 2021. Seals were captured using
nets and bio-loggers were attached during studies based in coastal communities
or on research ships. Ship-based capture events occurred during spring near the
southern ice edge in the Bering Sea between 2005 and 2018. Land-based capture
events occurred between 2005 and 2020 from May to October, generally between the
coastal communities of Scammon Bay, Alaska in the Bering Sea, Utqiaġvik, Alaska
in the Chukchi Sea, and Nuiqsut in the Beaufort Sea. Data from additional
deployments along the Kamchatka peninsula in the western Bering Sea are also
included. We refer readers to Figure \@ref(fig:deployplot) and the primary literature
for detailed capture and bio-logger attachment methods (see also Supplemental
Material, [S1](#s1)).

(ref:deployMapCaption) **Initial bio-logger deployment areas for each seal species (bearded, ribbon, and spotted seals) in the study between 2005 and 2020 across multiple regions within the Bering, Chukchi, and western Beaufort seas.** \linebreak Solid regions shown for each species are minimum concave polygons buffered by 60 km for enhanced visibility. The larger, shaded region indicates the spatial extent of aerial survey effort to date in the Bering, Chukchi, and Beaufort seas. Deployments were initiated across a range of months but only data from 1 March to 15 July were included in the analysis. See Figure \@ref(fig:dataMap) for the spatial distribution of observed data used in the study and Supplemental Materials \@ref(tab:deploymentTable) for additional deployment details.

@@

We subset haul-out behavior data from 
`r species_n %>% .$n %>% sum()` bio-loggers deployed on 
`r species_n %>% filter(species == "Bearded seal") %>% .$n` bearded, 
`r species_n %>% filter(species == "Ribbon seal") %>% .$n` ribbon, and 
`r species_n %>% filter(species == "Spotted seal") %>% .$n` spotted seals 
to include only records from 1 March to 15 July between 
`r min(lubridate::year(analysis_data$haulout_dt))` and 
`r max(lubridate::year(analysis_data$haulout_dt))`. Bio-loggers were of the
'SPLASH' or 'SPOT' family of tags developed by Wildlife Computers (Redmond,
Washington, USA). Deployments consisted of either a single 'SPLASH' device, a
single 'SPOT' device, or both types. Devices were either adhered to the hair
on the seal or attached through the rear flipper inter-digital webbing. The use
of bio-loggers adhered to the back or head provides some benefits over flipper
mounted devices (e.g. increased satellite transmittal rates, locations at sea)
but these are lost during the following annual molt, which, depending on
deployment date, limits the duration of haul-out data they provide.
Additionally, bio-loggers attached to the head or dorsal region are often dry
while the seal is floating at the surface, inducing a slight positive bias in
the hourly percent-dry values reported by the bio-logger. For this study, in
cases where both bio-logger types were deployed, we preferred hourly percent-dry
observations from the flipper tag. All data were transmitted by the deployed
instruments via the Argos satellite network and location data were either
derived from Argos transmissions or transmitted FastLoc GPS data.

Sex as well as age class (non-dependent _young-of-the-year_, sexually immature
_subadults_, and mature _adults_) were estimated at the time of deployment by
various combinations of length, claw growth ridges [@mclaren1958a], and pelage
characteristics. Seals determined to be less than one year were
classified as young-of-the-year. For those bio-loggers deployed on
young-of-the-year and transmitting into the next year, the age class was
advanced to subadult on 1 March of the following year. Subadults are those seals
likely greater than one year of age but less than four years. Adults are
individuals that are likely older than four years. Table \@ref(tab:tblDeploy)
provides a summary of these deployments and data received from them.

@@

Tags that fall off due to molt, attachment failure, or seal mortality and remain
on ice or land may continue to send data to satellites; i.e., a detached
bio-logger that is dry (either on ice or land) will record and transmit data
suggesting the seal is hauled out. Therefore, end times of each deployment were
identified by examining bio-logger locations, percent-dry records, and dive
behavior (if available) to determine when bio-loggers ceased providing data
consistent with seal behavior. For example, a data record that ends with several
consecutive days (~10+ days) of 100% dry observations and with locations
indicating the tag was on land would be truncated to not include the final
stretch of 100% dry observations. The vast majority of deployments end with the
device detaching in the water and the deployment end date is obvious. There is
no perfect algorithm for identifying deployment end dates and each deployment in
question must be considered separately. While not perfect, we are confident our
reliance on expert opinion and examination of multiple data streams provides the
best option. Data outside of the deployment start and end times were discarded
prior to analysis.

(ref:examplePlotCaption) **Example percent-dry actogram from bio-logger data**
\linebreak Haul-out behavior observations recorded by a bio-logger deployed on a
ribbon seal over two years during the months of March, April, and May. This
actogram plot represents the transmitted hourly percent-dry values. Not all data
during a deployment are reliably transmitted from the bio-logger and since data
are transmitted in 24-hour chunks occasional missing days are present in the
record. Here, missing data that were not successfully recieved from the device
are represented as empty rectangles.

@@

In addition to sex and age class, we analyzed variables that might help explain
variation in haul-out probabilities. These included day-of-year (for seasonal
effects) and local solar hour (for diurnal effects). We calculated local solar
hour using the {solaR} package [@perpinan2012] within the R statistical
environment [@rcoreteam2021] based on the weighted daily average locations. We
also linked the weighted average daily locations to weather values from the
North American Regional Reanalysis (NARR) model produced by the National Centers
for Environmental Prediction [@mesinger2006a]. The NARR model assimilates
observational data to produce a long-term picture of weather over North America
and portions of the surrounding seas. Weather variables are made available
across the region 8 times daily. For this study, NARR weather values were subset
to the extent of our study area over the Bering, Chukchi, and Beaufort seas at 3-hr
intervals based on the native grid resolution of 32 km (1024 km^2^). The
following meteorological variables are known to affect haul-out behavior in
other Arctic pinnipeds [@reder2003; @udevitz2009; @perry2017] and were
interpolated and assigned to daily seal locations using a bilinear method: 1)
air temperature at 2 m above the Earth's surface, 2) wind consisting of
northerly and easterly vector components converted to wind speed using the
Euclidean norm, 3) barometric pressure at sea level, and 4) precipitation 
(Table \@ref(tab:tblCovariates)).

@@

We did assess whether the annual variation in maximum spring sea ice extent in the
Bering Sea influenced the seasonal peak of seal haul-out probability. In
particular, we used sea ice concentration data from the Nimbus-7 SMMR and DMSP
SSM/I-SSMIS Passive Microwave Dataset, Version 1 [@cavalieri1996] to calculate
maximum sea ice extent. All sea ice concentration grid cells (25 km^2^) in the
study area with greater than 15% concentration were counted daily to get the total sea
ice extent for each day between 15 February and 15 July across all years.
Maximum spring sea ice extent was simply the largest daily count of grid cells
with greater than 15% concentration for each year.

@@

Haul-out records for seals are often characterized by sequential hours spent
basking on ice alternating with long periods in the water (Figure
\@ref(fig:examplePlot)). Commonly used statistical models for binary data (e.g.
logistic regression) assume independence among responses, an assumption that is
clearly violated if hourly responses are modeled. Any analysis that ignores
temporal autocorrelation in responses will thus have overstated precision
[@betts2006].

@@

To properly account for temporal dependence and to take advantage of computational
efficiency, we used generalized linear mixed pseudo-models (GLMPMs;
[@verhoef2010a]) to model variation in haul-out behavior as a function of (1)
covariate predictors, (2) temporally autocorrelated random effects, and (3)
individual random effects representing heterogeneity in individual behavior. We
used the glmmLDTS package [@verhoef2010a] to implement GLMPMs. We explored two
different model formulations for our data and we fit separate models to bearded, 
ribbon, and spotted seal data sets as we expected differing behavior by species. 
Separate models for each species were also
needed because a single, very large data set proved computationally intractable.
In our first model formulation and for each species, we fitted a year-independent
model that predicted average haul-out behavior as a function of demographic,
environmental, seasonal, and diurnal effects. Second, for ribbon and spotted
seals (which had considerably more data than bearded seals), we fitted models
that included all the effects from the first model, but also permitted annual
variation in haul-out timing. This second class of models was used to examine
whether haul-out patterns varied by year and to determine the annual timing of
apparent peaks in haul-out behavior. For both models, we assumed an hourly
Bernoulli response (i.e., whether tags were mostly dry or mostly wet) where
the linear predictor was modeled on the logit scale. This is consistent with
previous approaches [@verhoef2014a, @london2012] and only
`r (1 - (portion_10(analysis_data) + portion_90(analysis_data))) * 100`% of our
observations fell between 10% and 90% hourly percent-dry.

@@

Visualizing the marginal or conditional effect of an individual weather
covariate on haul-out probability was difficult in this analysis because of the
collinearity between covariates as well as the spatial and temporal variation
across such a large region. The relationship of each weather covariate with
haul-out probability, averaged over the other weather conditions, was more
variable than model coefficients would imply. That said,
important insights can be gained from plots of marginal effects. To create these
plots, we predicted haul-out probability across the full range
of weather covariate values while fixing hour of the day at local solar noon and
day-of-year at 15 May. The
visualizations also include vertical lines representing 95% confidence intervals
around the predicted haul-out probability to better communicate the variation in
model uncertainty.

@@

Figure \@ref(fig:dataMap) shows the spatial distribution of weighted locations
with available haul-out behavior data used for analysis of each species across
the study area. Figure \@ref(fig:dataCal) shows the temporal distribution of all
haul-out data across the study season for each species. Observations of ribbon
and spotted seals were concentrated in the months of May and June due to the
timing of deployment (April and May) and the timing of molt (May and June).
During molt, seals (and their attached bio-loggers) spend more time out of the
water and more data are transmitted. Molt timing also impacts when many
deployments end as any bio-loggers adhered to the hair will fall off. Relative
to the other species in the study, there were fewer deployments of bio-loggers
on bearded seals. This resulted in fewer data observations overall and
noticeably lower in numbers May and June. The majority of bearded seal
deployments started later in the summer and, by May, bio-loggers had either
fallen off or their batteries were depleted.

@@

(ref:mapCaption) **Spatial distribution of haul-out records from 1 March through July 15 for each of the three species.** \linebreak Linking location estimates with haul-out records in space and time allows for inclusion of weather covariates in the final model. For this visualization, data were collated across all years between 2005 and 2020 and each hexagonal cell represents an area of 50 km^2^

@@

(ref:dataDistributionCaption) **Seasonal density of haul-out behavior observations by species** \linebreak Distribution of hourly percent-dry bio-logger data from 1 March to 15 July for each species. Data are grouped by day-of-year and presented in seal-hours collated across all years between 2005 and 2020. Density values are represented as a color gradient and bar height. The higher density of data from May and June in ribbon and spotted seals coincides with peak molting when seals (and their attached bio-loggers) are more likely hauled out. Additionally, many bio-logger deployments started in April and May. The overall reduced quantity of observations from bearded seals is reflective of the lower number of bio-logger deployments in the study.

@@

Models omitting year effects suggested that day-of-year, solar hour, age-sex
class, temperature, and wind substantially influenced haul-out behavior of all
three species, with $F$ tests producing $p$-values less than 0.05 for variables
embodying these effects and/or their interactions. Haul-out probabilities
typically increased throughout March and April, reaching a peak in May and early
June before declining again. Diurnal patterns were present, with maximum
haul-out behavior centered around local solar noon.

@@

Age and sex class were not included in the model for bearded seals due to our
lower sample size for adult and young-of-year age classes. As such, results are
shown for all ages (Figure \@ref(fig:beardedHOCal); see also 
\@ref(fig:beardedPredSE)). Additionally, after
approximately 9 months, 7 devices deployed on the rear flipper of bearded seals
reported implausible hourly percent-dry data (100% dry for several weeks but
indicative of movement and increasing transmission rates (see @boveng2013)). All
data after the first instance of unrealistic values were censored from this
analysis. Overall, bearded seals were less likely to haul out and had a bi-modal
distribution of haul-out probability across the day. In addition to a peak
around local solar noon, the bearded seal model predicted additional haul-out
activity around local midnight. In concert with the lower magnitude of haul-out
probability, bearded seal haul-out behavior was also more protracted throughout
the spring season compared to ribbon and spotted seals.

@@

(ref:beardedHOCalCaption) **Bearded seal predicted haul-out probability** \linebreak Predicted hourly haul-out probability of bearded seals (all ages and sex classes) from 1 March to 30 June for all age and sex classes combined. Predictions in July and before 15 March were not included to avoid spurious model predictions at the edge of the data range. Note, our model predicts lower haul-out probability for bearded seals overall compared to ribbon and spotted seals and the color scales are not directly comparable.

>>>>

  pupping, breeding, molting, and resting. In the Arctic, many of these
  activities occur in spring (April -- June) as sea ice begins to melt and
  retreat northward. Rapid acceleration of climate change in Arctic ecosystems
  is therefore of concern as the quantity and quality of suitable habitat is
  forecast to decrease. Improved estimates of seal population abundance are
  needed to properly monitor the impacts of these changes over time. In this
  paper, we use hourly percent-dry data from satellite-linked bio-loggers
  deployed between 2005 and 2021 to quantify the proportion of seals hauled out
  on ice. This information is needed to accurately estimate abundance from
  aerial survey counts of ice-associated seals (i.e., to correct for the
  proportion of animals that are in the water while surveys are conducted). In
  addition to providing essential data for survey 'availability' calculations,
  our analysis also provides insights into the seasonal timing and environmental

@@

  snow lairs analysis of their 'availability' to surveys requires special
  consideration; therefore,they were not included in this analysis. Using
  generalized linear mixed pseudo-models to properly account for temporal
  autocorrelation, we fit models with covariates of interest (e.g., day-of-year,
  solar hour, age-sex class, wind speed, barometric pressure, temperature,
  precipitation) to examine their ability to explain variation in haul-out
  probability. We found evidence for strong diel and within-season patterns in
  haul-out behavior, as well as strong weather effects (particularly wind and
  temperature). In general, seals were more likely to haul out on ice in the
  middle of the day and when wind speed was low and temperatures were higher.
  Haul-out probability increased through March and April, peaking in May and
  early June before declining again. The timing and frequency of haul-out events
  also varied based on species and age-sex class. For ribbon and spotted seals,
  models with year effects were highly supported, indicating that the timing and
  magnitude of haul-out behavior varied among years. However, we did not find
  broad evidence that haul-out timing was linked to annual sea ice extent. Our
  analysis emphasizes the importance of accounting for seasonal and temporal
  variation in haul-out behavior, as well as associated environmental
  covariates, when interpreting the number of seals counted in aerial surveys.

@@

Statutory requirements (e.g., United States Marine Mammal Protection Act (MMPA),
United States Endangered Species Act (ESA)) for timely estimates of population
abundance and trend mean improved aerial survey effort is needed for these
species. Those survey efforts must also be paired with improved knowledge of
haul-out behavior to ensure appropriate survey design, robust methods, and
accurate estimates. Several studies have contributed estimates of the
distribution and abundance of ice-associated seal species in the Arctic using
aerial surveys (e.g.,
@bengtson2005a, @conn2014a, and @verhoef2014a) and more recent efforts have
significantly expanded on previous survey effort. Such abundance studies are
conducted over very large areas and estimation of absolute abundance requires
making inference about numerous issues affecting the observation of seals on
ice. These include availability (only seals on ice are available to be counted),
detection probability (observers or automated detection systems may miss some
seals on ice), species misclassification, and possible disturbance of seals by
aircraft [@conn2014a; @verhoef2014a]. Refining these inferences will improve the
accuracy of abundance estimates and, hopefully, allow credible predictions about
the effects of climate disruptions on the abundance and distribution of Arctic
seal populations.

How ice-associated seals use sea ice as a haul-out platform varies between
species. Ribbon seals (_Histriophoca fasciata_) haul out of the water almost
exclusively on sea ice and are mostly pelagic outside the spring pupping,
breeding, and molting season [@boveng2018a]. Although spotted (_Phoca largha_)
and bearded (_Erignathus barbatus_) seals occasionally rest on coastal land,
they primarily use sea ice as a haul-out platform during the spring and early
summer [@frost2018]. Ringed seals (_Phoca hispida_) --- not included in this
study --- haul out on sea-ice but primarily use snow lairs on sea-ice during
winter and spring.

@@

In this study, we used 16 years of bio-logging data to investigate the haul-out
behavior of bearded, ribbon, and spotted seals in the Bering and Chukchi seas.
Our goals were threefold. First, we wished to establish baseline estimates for
the chronology of haul-out behavior in the critical spring season for each
species across different age and sex classes. Second, we sought to refine
estimates of haul-out availability corrections for aerial surveys in order to
improve estimates of seal abundance. Previously estimated availability
correction factors (e.g., @bengtson2005a, @conn2014a, and @verhoef2014a)
accounted for variables such as the time-of-day and day-of-year, but did not
investigate the impact of weather variables. Such variables have been
shown to influence walrus haul-out behavior [@udevitz2009] and we expect weather
conditions to also influence seal haul-out behavior and including them within
the model framework will benefit our estimates of seal availability during
aerial surveys. Third, we aimed to assess the annual variability in haul-out
timing and possible linkage to changes in the extent of seasonal sea ice between
2005 and 2020. Our work extends the scope of previous haul-out analyses,
includes the influence of weather variability, and investigates the
potential impact of changing sea-ice extent on the behavior of these species.

@@

For this study we used haul-out behavior data and location estimates from
bio-loggers deployed on bearded, ribbon, and spotted seals in the Bering,
Chukchi, and western Beaufort seas by multiple organizations as part of
collaborative investigations from 2005 through 2021. Seals were captured using
nets and bio-loggers were attached during studies based in coastal communities
or on research ships. Ship-based capture events occurred during spring near the
southern ice edge in the Bering Sea between 2005 and 2018. Land-based capture
events occurred between 2005 and 2020 from May to October, generally between the
coastal communities of Scammon Bay, Alaska in the Bering Sea, Utqiaġvik, Alaska
in the Chukchi Sea, and Nuiqsut in the Beaufort Sea (Supplemental
Material, [S1](#s1)). Data from additional
deployments along the Kamchatka peninsula in the western Bering Sea are also
included. We refer readers to Figure \@ref(fig:deployplot) and the primary literature
for detailed capture and bio-logger attachment methods (see also Supplemental
Material, [S1](#s1)).

(ref:deployMapCaption) **Initial bio-logger deployment areas for bearded, ribbon, and spotted seals in the study between 2005 and 2020 across multiple regions within the Bering, Chukchi, and western Beaufort seas.** \linebreak Solid regions shown for each species are minimum concave polygons buffered by 60 km for enhanced visibility. The larger, shaded region indicates the spatial extent of aerial survey effort to date in the Bering, Chukchi, and Beaufort seas. Deployments were initiated across a range of months but only haul-out data from 1 March to 15 July were included in the analysis. See Figure \@ref(fig:dataMap) for the spatial distribution of observed data used in the study and Supplemental Materials \@ref(tab:deploymentTable) for additional deployment details.

@@

We created a subset of haul-out behavior data from 
`r species_n %>% .$n %>% sum()` bio-loggers deployed on 
`r species_n %>% filter(species == "Bearded seal") %>% .$n` bearded, 
`r species_n %>% filter(species == "Ribbon seal") %>% .$n` ribbon, and 
`r species_n %>% filter(species == "Spotted seal") %>% .$n` spotted seals 
to include only records from 1 March to 15 July between 
`r min(lubridate::year(analysis_data$haulout_dt))` and 
`r max(lubridate::year(analysis_data$haulout_dt))`. Bio-loggers were of the
'SPLASH' or 'SPOT' family of tags developed by Wildlife Computers (Redmond,
Washington, USA). Deployments consisted of either a single 'SPLASH' device, a
single 'SPOT' device, or both types. Devices were either adhered to the hair on
the seal or attached through the rear flipper inter-digital webbing. The use of
bio-loggers adhered to the back or head provides some benefits over flipper
mounted devices (e.g. increased satellite transmittal rates, locations at sea)
but these fall off during the following annual molt, which, depending on
deployment date, limits the duration of haul-out data they provide especially
during the focus months of our study. Additionally, bio-loggers attached to the
head or dorsal region are often dry while the seal is floating at the surface,
inducing a slight positive bias in the hourly percent-dry values reported by the
bio-logger. For this study, in cases where both bio-logger types were deployed,
we preferred hourly percent-dry observations from the flipper tag. All data were
transmitted by the deployed instruments via the Argos satellite network and
location data were either derived from Argos transmissions or transmitted
FastLoc GPS data.

Sex and age class (non-dependent _young-of-the-year_, sexually immature
_subadults_, and mature _adults_) were estimated at the time of deployment by
various combinations of length, claw growth ridges [@mclaren1958a], and pelage
characteristics. Seals determined to be less than one year were classified as
young-of-the-year. For those bio-loggers deployed on young-of-the-year and
transmitting into the next year, the age class was advanced to subadult on 1
March of the following year -- the assumed anniversary of their birth. Subadults
are those seals likely greater than one year of age but less than four years.
Adults are individuals that are likely older than four years. Table
\@ref(tab:tblDeploy) provides a summary of these deployments and data received
from them.

@@

Tags that fell off due to molt, attachment failure, or seal mortality and remained
on ice or land may have continued to send data to satellites; i.e., a detached
bio-logger that is dry (either on ice or land) will record and transmit data
suggesting the seal is hauled out. Therefore, end times of each deployment were
identified by examining bio-logger locations, percent-dry records, and dive
behavior (if available) to determine when bio-loggers ceased providing data
consistent with seal behavior. For example, a data record that ended with several
consecutive days (~10+ days) of 100% dry observations and with locations
indicating the tag was on land were truncated to the final
stretch of 100% dry observations. The vast majority of deployments ended with the
device detaching in the water and the deployment end date was obvious. There is
no perfect algorithm for identifying deployment end dates and each deployment in
question was considered separately. While not perfect, we are confident our
reliance on expert opinion and examination of multiple data streams provided the
best option. Data outside of the deployment start and end times were discarded
prior to analysis.

(ref:examplePlotCaption) **Example percent-dry actogram from bio-logger data**
\linebreak Haul-out behavior observations recorded by a bio-logger deployed on a
ribbon seal over two years during the months of March, April, and May. This
actogram plot represents the transmitted hourly percent-dry values. Not all data
during a deployment are reliably transmitted from the bio-logger and since data
are transmitted in 24-hour chunks occasional missing days are present in the
record. Here, missing data that were not successfully received from the device
are represented as empty rectangles.

@@

In addition to sex and age class, we analyzed variables that might help explain
variation in haul-out probabilities. These included day-of-year (for seasonal
effects) and local solar hour (for diurnal effects). We calculated local solar
hour using the {solaR} package [@perpinan2012] within the R statistical
environment [@rcoreteam2021] based on the weighted daily average locations. We
also linked the weighted average daily locations to weather values from the
North American Regional Reanalysis (NARR) model produced by the National Centers
for Environmental Prediction [@mesinger2006a]. The NARR model assimilates
observational data to produce a long-term picture of weather over North America
and portions of the surrounding seas. Weather variables are made available
across the region 8 times daily. For this study, NARR weather values were limited
to the extent of our study area over the Bering, Chukchi, and Beaufort seas at 3-hr
intervals based on the native grid resolution of 32 km (1024 km^2^). The
following weather variables are known to affect haul-out behavior in
other Arctic pinnipeds [@reder2003; @udevitz2009; @perry2017] and were
interpolated and assigned to daily seal locations using a bilinear method: 1)
air temperature at 2 m above the Earth's surface, 2) wind consisting of
northerly and easterly vector components converted to wind speed using the
Euclidean norm, 3) barometric pressure at sea level, and 4) precipitation 
(Table \@ref(tab:tblCovariates)).

@@

@@

Haul-out records for seals are often characterized by sequential hours spent
hauled out on ice alternating with long periods in the water (Figure
\@ref(fig:examplePlot)). Commonly used statistical models for binary data (e.g.
logistic regression) assume independence among responses, an assumption that is
clearly violated if hourly responses are modeled. Any analysis that ignores
temporal autocorrelation in responses will thus have overstated precision
[@betts2006].

@@

To properly account for temporal dependence and to take advantage of
computational efficiency, we used generalized linear mixed pseudo-models
(GLMPMs; [@verhoef2010a]) to model variation in haul-out behavior as a function
of (1) covariate predictors, (2) temporally autocorrelated random effects, and
(3) individual random effects representing heterogeneity in individual behavior.
We used the glmmLDTS package [@verhoef2010a] to implement GLMPMs. We explored
two different model formulations for our data and we fit separate models to
bearded, ribbon, and spotted seal data sets as we expected differing behavior by
species. Separate models for each species were also needed because a single,
very large data set proved computationally intractable. In our first model
formulation and for each species, we fitted a year-independent model that
predicted average haul-out behavior as a function of demographic, weather,
seasonal, and diurnal effects. Second, for ribbon and spotted seals (which had
considerably more data than bearded seals), we fitted models that included all
the effects from the first model, but also permitted annual variation in
haul-out timing. This second set of models was used to examine whether
haul-out patterns varied by year and to determine the annual timing of apparent
peaks in haul-out behavior. For both models, we assumed an hourly Bernoulli
response (a binary 0-1 response dependent upon whether the tag was greater than 
50% dry for a given hour) where the linear predictor was 
modeled on the logit scale. This is
consistent with previous approaches [@verhoef2014a, @london2012] and only
`r (1 - (portion_10(analysis_data) + portion_90(analysis_data))) * 100`% of our
observations fell between 10% and 90% hourly percent-dry.

@@

Visualizing the marginal or conditional effect of an individual weather
covariate (where all other weather covariates are being held at mean values) on
haul-out probability was difficult in this analysis because of the collinearity
between covariates as well as the spatial and temporal variation across such a
large region. The relationship of each weather covariate with haul-out
probability, averaged over the other weather conditions, was more variable than
model coefficients would imply. That said, important insights can be gained from
plots of marginal effects. To create these plots, we predicted haul-out
probability across the full range of each weather covariate while fixing hour
of the day at local solar noon and day-of-year at 15 May. For the other weather
covariate values, we chose not to use a fixed mean value because we expect weather
to vary within day over the season (e.g. the temperature at solar noon will
gradually increase from March through June). To account for this, we fit a simple generalized
additive model for each weather covariate with smooth terms for day-of-year and 
solar hour. We used predicted values from the generalized additive model in lieu 
of holding other weather covariates at a fixed mean value which would not capture
seasonal change. The visualizations
also include vertical lines representing 95% confidence intervals around the
predicted haul-out probability to better communicate the variation in model
uncertainty.

We also assessed whether the annual variation in maximum spring sea ice extent in the
Bering Sea influenced the seasonal peak of seal haul-out probability. In
particular, we used sea ice concentration data from the Nimbus-7 SMMR and DMSP
SSM/I-SSMIS Passive Microwave Dataset, Version 1 [@cavalieri1996] to calculate
maximum sea ice extent. All sea ice concentration grid cells (25 km^2^) in the
study area with greater than 15% concentration were counted daily to get the total sea
ice extent for each day between 15 February and 15 July across all years.
Maximum spring sea ice extent was simply the largest daily count of grid cells
with greater than 15% concentration for each year. A separate regression model,
built on the results of the haul-out model, was used to evaluate the relationship
between the annual computed peak haul-out day (as the response) with the
maximum sea ice extent (as the predictor).

@@

Figure \@ref(fig:dataMap) shows the spatial distribution of weighted locations
with available haul-out behavior data used for analysis of each species across
the study area. Figure \@ref(fig:dataCal) shows the temporal distribution of all
haul-out data across the study season for each species. Observations of ribbon
and spotted seals were concentrated in the months of May and June due to the
timing of deployment (April and May) and the timing of molt (May and June).
During molt, seals (and their attached bio-loggers) spend more time out of the
water and more data are transmitted. Molt timing also impacts when many
deployments end as any bio-loggers adhered to the hair will fall off. Relative
to the other species in the study, there were fewer deployments of bio-loggers
on bearded seals. This resulted in fewer data observations overall and
noticeably lower in numbers May and June. The majority of deployments on bearded 
seals occurred in August and September and, by May, bio-loggers had either
fallen off or their batteries were depleted. Therefore, observations for bearded 
seals were concentrated in March (Figure \@ref(fig:dataCal)).

@@

(ref:mapCaption) **Spatial distribution of haul-out records from 1 March through July 15 for bearded, ribbon, and spotted seal.** \linebreak Linking location estimates with haul-out records in space and time allows for inclusion of weather covariates in the final model. For this visualization, data were collated across all years between 2005 and 2020 and each hexagonal cell represents an area of 50 km^2^

@@

(ref:dataDistributionCaption) **Seasonal distribution of haul-out behavior observations by species** \linebreak Distribution of hourly percent-dry bio-logger data from 1 March to 15 July for each species. Data are grouped by day-of-year and presented in seal-hours collated across all years between 2005 and 2020. Seal-hours of data are represented as a color gradient and bar height. The higher amount of data from May and June in ribbon and spotted seals coincides with peak molting when seals (and their attached bio-loggers) are more likely hauled out. Additionally, many bio-logger deployments started in April and May. The overall reduced quantity of observations from bearded seals is reflective of the lower number of bio-logger deployments in the study and the fact that the majority of deployments on bearded seals occurred in August and September; therefore most bio-loggers deployed had exhausted their batteries.

@@

Across all three seal species, generally, models omitting year effects suggested
that day-of-year, solar hour, age-sex class, temperature, and wind substantially
influenced haul-out behavior of all three species, with $F$ tests producing
$p$-values less than 0.05 for variables embodying these effects and/or their
interactions. Haul-out probabilities typically increased throughout March and
April, reaching a peak in May and early June before declining again. Diurnal
patterns were present, with maximum haul-out behavior centered around local
solar noon.

@@

Age and sex class were not included in the model for bearded seals due to our
lower sample size for adult and young-of-year age classes. As such, results are
shown for all ages (Figure \@ref(fig:beardedHOCal); see also
\@ref(fig:beardedPredSE)). Additionally, after approximately 9 months, 7 devices
deployed on the rear flipper of bearded seals reported implausible hourly
percent-dry data (100% dry for several weeks but indicative of movement and
increasing transmission rates (see @boveng2013)). All data after the first
instance of unrealistic values were censored from this analysis. In addition to
a peak around local solar noon, the bearded seal model predicted additional
haul-out activity around local midnight. In concert with the lower magnitude of
haul-out probability, bearded seal haul-out behavior was also more protracted
throughout the spring season compared to ribbon (Figure \@ref(fig:ribbonHOCal))
and spotted seals (Figure \@ref(fig:spottedHOCal)). Overall, bearded seals were
less likely to haul out and had a bi-modal distribution of haul-out probability
across the day.

@@

(ref:beardedHOCalCaption) **Bearded seal predicted haul-out probability** \linebreak Predicted hourly haul-out probability of bearded seals (all ages and sex classes) from 15 March to 30 June for all age and sex classes combined. Predictions in July and before 15 March were not included to avoid spurious model predictions at the edge of the data range. Weather covariate values in the prediction were based on a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season. Note, our model predicts lower haul-out probability for bearded seals overall compared to ribbon and spotted seals and the color scales are not directly comparable.
